---
title: "Current Conditions revamp code demo"
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-depth: 3
---

This document outlines a workflow for generating “current conditions” visuals for streamflow and groundwater sites. 
The goal is a reproducible R/Python pipeline that produces graphics showing present-day conditions in the context of historical percentiles (daily, weekly, or monthly — TBD).  
Feedback is welcome — especially around API limitations and efficient data-pull strategies.

## High-level skeleton

We break the workflow into **four independent processes**. Each can run on its own schedule. GitLab CI is the likely orchestration layer; AWS S3 is the storage backend.

1. **Site List Construction**: Build and store the authoritative list of “active” sites and their metadata.

2. **Regular Data Pull**: For each active site, collect the newest observations + percentile values, and assign categories (e.g., “below 10th percentile”).

3. **Visual Creation** Combine site metadata + categorized conditions into map frames.

4. **Display** Host/export visuals from a public S3 bucket.

Each section below expands on these stages.

## 1. Site List Construction

This job defines the set of sites tracked by the downstream pipeline. 
All downstream tasks assume this set is stable until the next rebuild.

### What counts as “active”?

Possible criteria:

- **Parameter codes:** must collect at least one pcode of interest (e.g., 00065, 72019).
- **Recency:** at least one observation in the last *X* months.
- **Period of record (POR):** must have ≥ *Y* years of data.  
  (E.g., deciles need ≥ 11 years; quartiles need ≥ 5.)
- **Frequency:** each time-grouping used by the visual must have ≥ *Y* observations across years.

### Alternate definition: use `/statistics` availability

Instead of building our own definition, we could simply use whatever sites are available through the `/statistics` endpoint.  

Pros:

- Very simple to check.
- We want to use `/statistics` for percentiles anyway, so checking availability doubles as pulling the percentiles we need.

Cons:

- `/statistics` excludes provisional data.
- Its POR/frequency rules may not match our needs, and exclude more sites that we want.
- Need to assess performance at large scales.

### Tentative Procedure

1. **Pull candidate IDs** via `/time-series-metadata`, filtering by:  

   - parameter_codes  
   - begin date ≥ POR threshold  
   - end date ≤ recency threshold  
   - *Note*: this initial pull only screens *possible* IDs; it does not check observation density.

2. **Split IDs** by statistic ID:

   - `00003`: daily mean  
   - `00011`: instantaneous (continuous)  
   - Everything else (“min”, “max”, etc.) is ignored but worth tracking for diagnostics.

3. **Branching decision: do we enforce observation frequency?**

#### If *not* enforcing frequency

Keep one time-series ID per site:

- Prefer daily mean (`00003`) when available.
- Choose the ID with the longest POR.
- Break ties using preferred pcodes (00060/00065 or 72019). Otherwise, pick an arbitrary timeseries.

#### If *enforcing* frequency

For each daily-mean ID:

1. Pull full POR from `/daily`.  
   (Groundwater additionally merges `/field-measurements`.)
2. Run `determine_data_sufficiency()` (see below).

For instantaneous IDs:

1. Skip if the site already has a sufficient daily-mean timeseries.  
2. Otherwise:
   - Pull full POR from `/continuous`.  
   - If identified earlier as “potentially salvageable,” merge with daily/field-measurement data.
   - Run `determine_data_sufficiency()` (see below).

Finally, for each monitoring location:

- Keep all timeseries deemed **sufficient**.  
- If multiple pcodes are sufficient for a specific site, choose the one with longest POR (tie-break as above).

### Pseudocode: `determine_data_sufficiency()`

Purpose:  
Determine whether a timeseries has enough historical coverage to compute percentiles for every “frame” of the visual.

Inputs:

- Observations from one endpoint (or merged sources).
- `F` = frame size (daily/weekly/monthly).
- `n_F` = number of frames in a year (e.g., 12 for monthly).  
- `Y` = required number of years of coverage (e.g., 11 for deciles).

Logic:

1. Group all observations into their corresponding frame group `G_i`.  
2. For each `G_i`:
   - Count number of years with ≥1 observation (call this `N_i`).  
3. Classify:
   - If all `N_i >= Y` → **sufficient**.  
   - If some `N_i < Y`:
     - If continuous data already in merged source → **insufficient**.  
     - If a corresponding 00011 ID exists → **potentially sufficient** (gaps may be filled).  
     - Else → **insufficient**.

Output: one of {sufficient, potentially_sufficient, insufficient} per timeseries ID.



## 2. Regular Data Pull

Executed daily/weekly/monthly depending on parameter type.

Steps:

1. Load active site list from S3.
2. For each site:
   - Pull most recent observation (daily or continuous).
   - Pull percentile values.
   - Categorize observation into percentile bins.
3. Merge all site categorizations into one table.
4. Push the table to S3.

Settings developers can tune:

- Frequency of the job.
- Fallback logic (e.g., use `/continuous` when `/daily` missing).
- Percentile resolution (quartiles, deciles, etc.).
- Whether frequency-based sufficiency is re-evaluated periodically.



## 3. Visual Creation

Inputs:

- Categorizations + metadata from S3.

Outputs:

- Map-like frames (e.g., PNG/SVG).
- Possibly animation slices (daily/weekly/monthly).

Configurable:

- Frame duration (`F`): daily/weekly/monthly.
- Output resolution and file types.
- Whether to generate animated sequences vs static tiles.

## 4. Display Layer

Steps:

1. Pull completed visuals from S3.
2. Make them available to downstream apps (public S3, CloudFront, etc.).

## Open questions / considerations

**Vizualization Questions**:

* What is the frame duration?
* What percentiles are needed?
* How often will the visual be updated?

**Streamflow SME Questions:**

* Is relationship between discharge and gage height *monotonic*? If yes, then we can pull both and compute percentiles without worrying about comparability.
* Any other pcodes we should consider than 00060, 00065?

**\statistics API questions**:

* What data are available through the service (e.g., only specific site types/parameter codes allowed/statistic IDs)?
* What criteria are used for computing day/month-of-year percentiles (e.g., POR, missing data handling, etc.)?

## Code demo: Initial filter with time-series-metadata

The `/time-series-metadata` endpoint includes filterable fields `parameter_codes`, `begin`, and `end`, which would allow us to quickly filter sites based on its POR and collected pcodes.
However, this doesn't take into account observation **frequency**.

A **timeseries ID** uniquely identifies a combination of:

- monitoring location
- pcode collected at that monitoring location
- statistic ID (e.g., daily mean, instantaneous)
- period of record (e.g., if a sensor is moved at a location, then a new timeseries ID is typically created)
- (in rare instances) additional metadata about the data collection. For example, a monitoring location might have two sensors that collect the exact same data but in different locations (e.g, on both sides of a sluice gate). These would have different timeseries IDs.

::: {.panel-tabset group="sitetype"}

Note that the `geometry` column returned by the API contains spatial point data for each row.
We'll be pulling site metadata like this later on, so we can set `skip_geometry` / `skipGeometry` to cut down on the response size (saves us memory, saves the API folks money).
We could set the `properties` argument to only return specific columns to further save on cost.

## Streamflow

```{python}
#| label: sf-setup
#| eval: true
from dataretrieval import waterdata
import os
import datetime as dt
import pandas as pd
import matplotlib.pyplot as plt

# ignore this:
envvar = open(".Renviron").read().split("\n")[0].replace("API_USGS_PAT1=", "")
os.environ["API_USGS_PAT"] = envvar

sf_pcodes = ['00060', '00065']

# get_ogc_data is low-level function for querying services
sf_pcode_meta = waterdata.utils.get_ogc_data(
  args = {'pcodes': sf_pcodes}, 
  output_id = "pcodes", 
  service = "parameter-codes")[0]

# only want sites with at least one observation in last X days. Used for initial
# filtering
earliest_date_needs_observation = dt.date.today() - pd.DateOffset(days = 365)

# sites need to have at least Y-year POR. Used for initial filtering
latest_date_por = dt.date.today() - pd.DateOffset(years = 10)

# Set this manually to reduce computational load. Earliest is '1700-01-01'
earliest_date_por = dt.date(1985, 1, 1)

# Span of time (in days) a single "frame" of the visual represents. Used when
# assessing if a site has sufficient data
update_timestep = 1

required_dates = pd.date_range(
  start = earliest_date_por, 
  end = dt.date.today(), 
  freq = str(update_timestep) + 'D'
  )
```

```{python}
#| label: sf-ts-ids
#| eval: true
# initial filter using parameter_code, begin, and end fields
sf_all_ts_ids = waterdata.get_time_series_metadata(
    parameter_code = sf_pcodes,
    begin = "1700-01-01T00:00:00Z/" + latest_date_por.strftime("%Y-%m-%dT%H:%M:%SZ"),
    end = earliest_date_needs_observation.strftime("%Y-%m-%dT%H:%M:%SZ") + "/..",
    skip_geometry = True,
)[0]

# See https://api.waterdata.usgs.gov/ogcapi/v0/collections/statistic-codes/items for other stat IDs
sf_all_ts_ids.groupby(["parameter_code", "statistic_id"]).size().sort_values(ascending = False)
```

```{python}
#| label: sf-ts-id-exploration
#| eval: true
sf_ts_ids_daily_mean = sf_all_ts_ids[sf_all_ts_ids.statistic_id == '00003']

sf_ts_ids_continuous = sf_all_ts_ids[sf_all_ts_ids.statistic_id == '00011']

sf_ts_ids_daily_mean.groupby('monitoring_location_id').size().sort_values(ascending = False)

# "sublocation_identifier" and "web_description" columns sometimes contain distinguishing information for sites with multiple TS IDs. 
sf_ts_ids_daily_mean[sf_ts_ids_daily_mean.monitoring_location_id=="USGS-411219096010601"][['monitoring_location_id', 'parameter_code','begin', 'end', 'sublocation_identifier', 'web_description']]
```

## Groundwater

I've created a few api.waterdata tokens and saved them to a local .Renviron file.
While I don't think you're *supposed* to make multiple API tokens, I do recommend saving your token to a .Renviron.
You can even call it `API_USGS_PAT=....` in the .Renviron file, saving you from needing to run the `Sys.setenv` call below.

```{r}
#| label: gw-setup
#| eval: true
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(purrr)
library(plotly)

Sys.setenv("API_USGS_PAT" = Sys.getenv("API_USGS_PAT2"))

# These are all some variation of "water level," meaning we can compute
# percentiles without worrying about their comparability with the other pcodes.
gw_pcodes <- c(
  "72019",
  "72150",
  "62600",
  "30210",
  "62611",
  "62610",
  "62613",
  "62612",
  "72231",
  "72232",
  "72230",
  "72229",
  "72227",
  "72228",
  "72226",
  "61055",
  "62601"
)

gw_pcode_meta <- read_waterdata_parameter_codes(parameter_code = gw_pcodes)

# NGWMN inclusion criteria:
## At least one observation in last 406 days
earliest_date_needs_observation <- Sys.Date() - lubridate::days(406)
## At least 10-year POR
latest_date_por <- Sys.Date() - lubridate::years(10)
## How long will a given "frame" of the visual represent? "day", "week",
## "month", etc. Used to determine whether a site has enough data to be included in the visual.
update_timestep <- "month"

# Earliest date to fetch data when computing percentiles -- set later to reduce
# comp. time. Earliest value is "1700-01-01"
earliest_date_por <- "1985-01-01"

# Compute timesteps for which an "active" site should have at least one
# observation in each
required_dates <- seq.Date(
  earliest_date_por,
  lubridate::floor_date(Sys.Date(), unit = update_timestep),
  by = update_timestep
)
```

```{r}
#| label: gw-ts-ids
#| eval: true
gw_all_ts_ids <- read_waterdata_ts_meta(
  begin = paste0("1700-01-01/", latest_date_por),
  end = paste0(earliest_date_needs_observation, "/.."),
  parameter_code = gw_pcodes,
  skipGeometry = TRUE
) |>
  select(
    monitoring_location_id,
    parameter_code,
    statistic_id,
    begin,
    end,
    time_series_id,
    everything()
  ) |>
  arrange(monitoring_location_id, parameter_code, statistic_id)

# 2. daily data are going to be the easiest to work with. We should first figure
# out which sites have enough daily data before looking at alternatives
ml_ids_daily_means <-
  gw_all_ts_ids |>
  filter(statistic_id == "00003")

# 3. some sites might not have daily data but *do* have continuous.
ml_ids_continuous <-
  gw_all_ts_ids |>
  filter(statistic_id == "00011")
```

:::

### Code demo: Determining observation frequency

Insufficient data could impact our ability to compute percentiles.
We might consider filtering sites with insufficient data when creating the active site list.

::: {.callout-important}

Whether a site has "sufficient" data is impacted by other decisions we make. For example, 

* Computing *quartiles* only requires 5 observations for a given frame duration over a site's POR.
* We could choose to have each frame represent one *week*'s worth of data, meaning a site would only need to have at least one observation in a 7-day period.
  * Said another way, only sites that have multiple periods of 7 days straight of missing data would be excluded from the visual.
* We could potentially use packages like `hyswap` to *interpolate* percentiles when there's insufficient data, though I'd want to check with an SME about the statistical validity/assumptions of these calculations.

:::

The cost of assessing observation frequency is high given the APIs available to us.
From my understanding, there is not (nor are the plans) for an endpoint that specifically summarizes observation counts for a given site.
Therefore, we would need to pull the full POR of every site (that passes the initial `/time-series-metadata` filters) to assess data sufficiency for a given list of percentiles and frame duration.

The `/daily` endpoint suffices for our use case and will be computationally easier to pull from than `/continuous`.
The `/field-measurements` endpoint will be helpful for pulling groundwater data, though most likely for filling gaps in `/daily` data since there are much longer spans between manually collected observations.
Candice also mentioned pulling groundwater data from `/samples-data`, though that service both seems to be slow (fewer resources) and does not provide much "extra" data.


::: {.panel-tabset group="sitetype"}

## Streamflow

I'm more dextrous with R than Python, so I'd recommend checking out the R code or the pseudocode I provided in the tentative active site list skeleton for guidance on assessing observation frequency.

## Groundwater

The code below is a start to implementing the pseudocode in the "Tentative active site list skeleton" section.
It doesn't implement the full logic of piecing together data from the three API endpoints (`/daily`, `/field-measurements`, `/continuous`), but it does do some data gap computation.

```{r}
#| label: gw-field-measurements
#| eval: false
# this cell isn't evaluated by-default as it takes a long time to run.
# It could be useful for filling data gaps in groundwater sites

# Pull full POR from field-measurements.
# First fetch sub-list of sites with at least one obs in last 406 days
gw_recent_field_measurements <-
  read_waterdata_field_measurements(
    parameter_code = gw_pcodes,
    time = paste0(earliest_date_por, "/..")
  )

gw_ml_ids_field_measurement <- unique(
  gw_recent_field_measurements$monitoring_location_id
)

# Among sites with at least one obs in last 406 days, determine which have a
# long enough POR
gw_old_field_measurements <-
  purrr::map_dfr(
    split(
      unique(gw_recent_field_measurements$monitoring_location_id),
      ceiling(
        seq_along(unique(gw_recent_field_measurements$monitoring_location_id)) /
          100
      )
    ),
    ~ read_waterdata_field_measurements(
      monitoring_location_id = .x,
      parameter_code = gw_pcodes,
      skipGeometry = TRUE
    )
  )

gw_field_measurements_sufficient_por <-
  gw_old_field_measurements |>
  sf::st_drop_geometry() |>
  filter(time <= latest_date_por) |>
  distinct(monitoring_location_id) |>
  pull(monitoring_location_id)

gw_old_field_measurements |>
  sf::st_drop_geometry() |>
  filter(monitoring_location_id %in% gw_field_measurements_sufficient_por) |>
  mutate(month = lubridate::floor_date(time, unit = "months")) |>
  select(monitoring_location_id, parameter_code, month) |>
  tidyr::nest(month)
```

```{r}
#| label: gw-assess-sufficiency
#| eval: true
# Assess frequency of observations -- is there at least one obs per timestep
# over POR?

# Fetch daily data since earliest_date_por -- requires splitting TS IDs into
# chunks of 50
all_daily_obs <-
  purrr::map_dfr(
    split(
      ml_ids_daily_means$time_series_id,
      ceiling(seq_along(ml_ids_daily_means$time_series_id) / 50)
    )[1], # NOTE: testing with one chunk of 50 IDs
    ~ read_waterdata_daily(
      time_series_id = .x,
      time = paste0(earliest_date_por, "/.."),
      skipGeometry = TRUE,
      limit = 50000
    )
  )

# Determine if there are any months of missing daily data for each monitoring
# location
daily_missing_timesteps <-
  all_daily_obs |>
  mutate(month = lubridate::floor_date(time, unit = update_timestep)) |>
  arrange(time_series_id, month) |>
  distinct(
    monitoring_location_id,
    parameter_code,
    statistic_id,
    time_series_id,
    month
  ) |>
  group_by(monitoring_location_id, parameter_code) |>
  summarize(
    missing_timesteps = list(required_dates[!required_dates %in% month])
  ) |>
  left_join(
    distinct(ml_ids_daily_means, monitoring_location_id, parameter_code, begin),
    by = c("monitoring_location_id", "parameter_code")
  ) |>
  # only want to track missing months since start of POR ("begin") for each site:
  mutate(
    missing_timesteps = purrr::map2(missing_timesteps, begin, ~ .x[.x >= .y])
  ) |>
  select(-begin) |>
  # only keep ML IDs with missing daily data
  filter(purrr::map_lgl(missing_timesteps, ~ length(.x) > 0))


# These are sites with no gaps in daily data for their entire POR -- these will
# be easiest to pull data for
gw_ml_ids_complete_daily <-
  all_daily_obs |>
  distinct(monitoring_location_id, parameter_code, time_series_id) |>
  anti_join(
    daily_missing_timesteps,
    by = c("monitoring_location_id", "parameter_code")
  )

# These are sites that have gaps in daily data that *may* be filled with
# continuous data, we'll check below
gw_ml_ids_partial_daily <-
  ml_ids_continuous |>
  select(monitoring_location_id, parameter_code, time_series_id, begin) |>
  right_join(
    daily_missing_timesteps,
    by = c("monitoring_location_id", "parameter_code")
  ) |>
  # check if missing daily data dates are before a site started collection
  # continuous data. If yes, then there's really no hope to fill the daily data
  # gaps.
  filter(purrr::map2_lgl(missing_timesteps, begin, ~ all(.x > .y)))


# These are sites with daily data gaps that can't be filled with continuous data
# (daily data gaps occur before continuous measurement began). We can't do
# anything with these. NOTE: maybe we should check /field-measurements data too
gw_ml_ids_lost_causes <- all_daily_obs |>
  distinct(monitoring_location_id, parameter_code, time_series_id) |>
  anti_join(
    bind_rows(gw_ml_ids_complete_daily, gw_ml_ids_partial_daily),
    by = c("monitoring_location_id", "parameter_code")
  )

gw_ml_ids_lost_causes
```


:::

### Code demo: Fetching site metadata

The `/monitoring-location` endpoint complains if you ask for too many monitoring location IDs at one.
I have an example in the Groundwater section that shows splitting and iterating over a list of monitoring location IDs, broken into chunks of 50 IDs.

::: {.panel-tabset group="sitetype"}

## Streamflow

```{python}
#| label: sf-site-metadata
#| eval: true
sf_site_metadata = waterdata.get_monitoring_locations(monitoring_location_id=sf_all_ts_ids.monitoring_location_id.unique().tolist()[0:9])

sf_site_metadata
```

## Groundwater

```{r}
#| label: gw-site-metadata
#| eval: true
gw_ml_ids <- unique(c(ml_ids_daily_means$monitoring_location_id, ml_ids_continuous$monitoring_location_id))

gw_ml_metadata <-
  purrr::map_dfr(
    split(gw_ml_ids, ceiling(seq_along(gw_ml_ids) / 100)),
    read_waterdata_monitoring_location
  ) |>
  left_join(
    bind_rows(ml_ids_daily_means, ml_ids_continuous) |>
      select(monitoring_location_id, parameter_code, statistic_id),
    by = "monitoring_location_id"
  )

# Plot ML IDs by parameter code
{
  gw_ml_metadata |>
    ggplot() +
    geom_sf(aes(colour = parameter_code))
} |>
  plotly::ggplotly()
```

:::

## Code demo: regular data pull

The [`/statistics`](https://api.waterdata.usgs.gov/statistics/v0/docs) API provides day and month-of-year statistics.
More experimentation is needed to determine the limitations of the `/statistics` API.
For example, I'm not sure what the logic is for computing percentiles -- what's the minimum POR required?
Are there constraints on what sites/parameter codes provided by the API?

Additionally, as you might be able to tell from the R Groundwater example code, the API response is densely nested JSON.
We'd want to write some Python/R helper functions to parse the output consistently.

::: {.panel-tabset group="sitetype"}

## Streamflow

```{python}
#| label: sf-statistics-example
import requests
import pandas as pd

# Build and send request
base_url = "https://api.waterdata.usgs.gov/statistics/v0/observationNormals"

params = {
    "monitoring_location_id": "USGS-07141780",
    "computation_type": "percentile",
}

resp = requests.get(base_url, params=params)
resp.raise_for_status()
data = resp.json()

# Extract values table (mirrors gw_stats$features$properties$data[[1]]$values[[1]])
values = data["features"][0]["properties"]["data"][0]["values"][0]

df = pd.DataFrame(values)

# Filter equivalent to dplyr::filter(time_of_year_type == "month_of_year")
df = df[df["time_of_year_type"] == "month_of_year"]

df
```

## Groundwater

```{r}
#| label: gw-statistics-example
gw_stats <-
  httr2::request("https://api.waterdata.usgs.gov") |>
  httr2::req_url_path("statistics/v0/observationNormals") |>
  httr2::req_url_query(
    "monitoring_location_id" = "USGS-430739078502701",
    "computation_type" = "percentile"
  ) |>
  httr2::req_perform() |>
  httr2::resp_body_string() |>
  jsonlite::fromJSON()

tibble::tibble(gw_stats$features$properties$data[[1]]$values[[1]]) |>
  filter(time_of_year_type == "month_of_year")
```

:::

## Computational Details

Note that many of the api.waterdata endpoints have implicit limits -- they'll often throw a 500 or 403 if they don't like something you've requested.
I've yet to determine whether these limitations derive from a request character limit, size, or some other factor.
I've found it safest to chunk large requests, which the helper functions below are meant to facilitate.


I have some R code below for setting up parallelization.
The `furrr` package has drop-in replacements of `purrr` functions that make it very easy to parallelize code; example: `purrr::map_dfr` -> `furrr::future_map_dfr`.

```{r}
#| label: gw-parallelization
#| eval: false
library(future)
library(furrr)
library(parallelly)
if (parallelly::supportsMulticore()) {
  future::plan(future::multicore,
    workers = ceiling(parallel::detectCores() / 3)
  )
} else {
  future::plan(future::multisession,
    workers = ceiling(parallel::detectCores() / 3)
  )
}

# Run this to "turn off" parallelization
# future::plan(future::sequential)
```

