---
title: "Current Conditions revamp code demo"
format: 
  html:
    embed-resources: true
    code-fold: show
---


::: {.panel-tabset}

## Streamflow

I'm more dextrous with R than Python, so this code is mainly intended as basic starting point for exploration.
All of this code runs without parallelization (I'm less familiar with how to do that in Python), so

```{python}
#| label: setup-py

from dataretrieval import waterdata
import os
import datetime as dt
import pandas as pd
import matplotlib.pyplot as plt

# ignore this:
envvar = open(".Renviron").read().split("\n")[2].replace("API_USGS_PAT3=", "")

os.environ["API_USGS_PAT"] = envvar
```

Discharge (00065) and gage height (00060) are the most popular streamflow pcodes.
We should still reach out to a Streamflow SME about other pcodes we should consider fetching (e.g., tidally filtered gage height).

```{python}
#| label: sf-table-setting

sf_pcodes = ['00060', '00065']

# get_ogc_data is low-level function for querying services
sf_pcode_meta = waterdata.utils.get_ogc_data(
  args = {'pcodes': sf_pcodes}, 
  output_id = "pcodes", 
  service = "parameter-codes")[0]

# only want sites with at least one obs. in last X days. Used for initial
# filtering
earliest_date_needs_observation = dt.date.today() - pd.DateOffset(days = 365)

# sites need to have at least Y-year POR. Used for initial filtering
latest_date_por = dt.date.today() - pd.DateOffset(years = 10)

# Set this manually to reduce computational load. Earliest is '1700-01-01'
earliest_date_por = dt.date(1985, 1, 1)

# Span of time (in days) a single "frame" of the visual represents. Used when
# assessing if a site has sufficient data
update_timestep = 1

required_dates = pd.date_range(
  start = earliest_date_por, 
  end = dt.date.today(), 
  freq = str(update_timestep) + 'D'
  )
```

A **timeseries ID** uniquely identifies a combination of:

- monitoring location
- pcode collected at that monitoring location
- statistic ID (e.g., daily mean, instantaneous)
- period of record (e.g., if a sensor is moved at a location, then a new timeseries ID is typically created)
- (in rare instances) additional metadata about the data collection. For example, a monitoring location might have two sensors that collect the exact same data but in different locations (e.g, on both sides of a sluice gate). These would have different timeseries IDs.

Note that the `geometry` column contains spatial point data for each row.
We'll be pulling site metadata like this later on, so we can set `skip_geometry` to cut down on the response size (saves us memory, saves the API folks money).
We could set the `properties` argument to further whittle the response down, if we want.

```{python}
#| label: sf-timeseries-ids

sf_all_ts_ids = waterdata.get_time_series_metadata(
    parameter_code = sf_pcodes,
    begin = "1700-01-01T00:00:00Z/" + latest_date_por.strftime("%Y-%m-%dT%H:%M:%SZ"),
    end = earliest_date_needs_observation.strftime("%Y-%m-%dT%H:%M:%SZ") + "/..",
    skip_geometry = True,
)[0]

# instant. (00011) discharge is most commonly reported, mean daily (00003) discharge being less common.
# instant. and daily mean gage height are about equally common
# See https://api.waterdata.usgs.gov/ogcapi/v0/collections/statistic-codes/items for other stat IDs
sf_all_ts_ids.groupby(["parameter_code", "statistic_id"]).size().sort_values(ascending = False)
```

Let's explore the timeseries IDs a bit more.
Sites with daily mean data will be easiest to work with, so we should first assess daily data coverage.
Only if a site has missing daily mean data (either periodically missing or 00003 simply isn't collected there) should we consider pulling from /continuous.

```{python}
#| label: sf-daily-mean-vs-instant
sf_ts_ids_daily_mean = sf_all_ts_ids[sf_all_ts_ids.statistic_id == '00003']

sf_ts_ids_continuous = sf_all_ts_ids[sf_all_ts_ids.statistic_id == '00011']

sf_ts_ids_daily_mean.groupby('monitoring_location_id').size().sort_values(ascending = False)
```

As an example of multiple timeseries IDs for one monitoring location, consider this site with 4 gage height TS IDs, each distinguished by their "sublocation_identifier". Sometimes the web_description column contains distinguishing information too. 

```{python}
#| label: sf-noteworthy-site
sf_ts_ids_daily_mean[sf_ts_ids_daily_mean.monitoring_location_id=="USGS-411219096010601"][['monitoring_location_id', 'parameter_code','begin', 'end', 'sublocation_identifier', 'web_description']]
```

I'm not sure how best to deal with these multiple timeseries.
We might consider filtering to the IDs with the longest period of record (`end - begin`).
If there are ties based on this ranking, we might have to pull all data for both time series, then determine which have fewer data gaps (e.g., observations missing for a given day).
At some point, we'll need to pick one timeseries ID for each monitoring location.

Here's an example of pulling all daily data for one site:

```{python}
#| label: sf-daily
sf_daily_obs = waterdata.get_daily(
  monitoring_location_id = sf_ts_ids_daily_mean.monitoring_location_id.unique().tolist()[0],
  skip_geometry = True
  )

sf_daily_obs
```

Note that many of the api.waterdata endpoints have implicit limits -- they'll often throw a 500 or 403 if they don't like something you've requested.
I've yet to determine whether these limitations derive from a request character limit, size, or some other factor.
I've found it safest to chunk large requests, which the helper functions below are meant to facilitate.

```{python}
#| label: sf-site-metadata
def chunked(iterable, size):
    """Yield successive `size`-sized chunks from iterable."""
    for i in range(0, len(iterable), size):
        yield iterable[i:i + size]

def fetch_monitoring_locations(ids, chunk_size=50):
    """
    Fetch metadata for a large list of monitoring location IDs using chunked requests.
    
    Parameters
    ----------
    ids : list of str
        Monitoring location IDs.
    chunk_size : int
        Number of IDs per API request (depends on service limits).
        
    Returns
    -------
    pandas.DataFrame
        Combined metadata table.
    """
    frames = []

    for chunk in chunked(ids, chunk_size):
        df, _ = waterdata.get_monitoring_locations(monitoring_location_id=chunk)
        frames.append(df)

    if frames:
        return pd.concat(frames, ignore_index=True)

    return pd.DataFrame()



sf_site_metadata = fetch_monitoring_locations(ids = sf_all_ts_ids.monitoring_location_id.unique().tolist(), chunk_size = 100)

sf_site_metadata.plot(markersize = 1)
plt.show()
```

## Groundwater

I've created a few api.waterdata tokens and saved them to a local .Renviron file.
While I don't think you're *supposed* to make multiple API tokens, I do recommend saving your token to a .Renviron.
You can even call it `API_USGS_PAT=....` in the .Renviron file, saving you from needing to run the `Sys.setenv` call below.

Note also that I've set up some parallelization code below.
I typically use a `multisession` plan for parallelizing, which fires up multiple R sessions in the background to perform computation on a subset of data you've provided.
However, `multicore` plans are more computationally efficient than `multisession`, though they only work on certain operating systems (including MacOS, I think?).
See the `gw-field-measurements` code chunk below for an example of using `furrr::future_map_dfr` to parallelize a `purrr::map_dfr` call.

```{r}
#| label: setup-r
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(purrr)
library(plotly)

Sys.setenv("API_USGS_PAT" = Sys.getenv("API_USGS_PAT3"))

# Uncomment for parallelization setup

# library(future)
# library(furrr)
# library(parallelly)
# if(parallelly::supportsMulticore()){
#   future::plan(future::multicore,
#                workers = ceiling(parallel::detectCores() / 3))
# } else{
#   future::plan(future::multisession,
#                workers = ceiling(parallel::detectCores() / 3))
# }

# Run this to "turn off" parallelization
# future::plan(future::sequential)
```

Candice sent us a big list of Groundwater pcodes to pull.
These are all some variation of "water level," meaning we can compute percentiles without worrying about their comparability with the other pcodes.
I also include the inclusion criteria used by NGWMN.

```{r}
#| label: gw-table-setting
gw_pcodes <- c(
  "72019",
  "72150",
  "62600",
  "30210",
  "62611",
  "62610",
  "62613",
  "62612",
  "72231",
  "72232",
  "72230",
  "72229",
  "72227",
  "72228",
  "72226",
  "61055",
  "62601"
)

gw_pcode_meta <- read_waterdata_parameter_codes(parameter_code = gw_pcodes)

# NGWMN inclusion criteria:
## At least one observation in last 406 days
earliest_date_needs_observation <- Sys.Date() - lubridate::days(406)
## At least 10-year POR
latest_date_por <- Sys.Date() - lubridate::years(10)
## How long will a given "frame" of the visual represent? "day", "week",
## "month", etc. Used to determine whether a site has enough data to be included in the visual.
update_timestep <- "month"

# Earliest date to fetch data when computing percentiles -- set later to reduce
# comp. time. Earliest value is "1700-01-01"
earliest_date_por <- "1985-01-01"

# Compute timesteps for which an "active" site should have at least one
# observation in each
required_dates <- seq.Date(
  earliest_date_por,
  lubridate::floor_date(Sys.Date(), unit = update_timestep),
  by = update_timestep
)
```

A **timeseries ID** uniquely identifies a combination of:

- monitoring location
- pcode collected at that monitoring location
- statistic ID (e.g., daily mean, instantaneous)
- period of record (e.g., if a sensor is moved at a location, then a new timeseries ID is typically created)
- (in rare instances) additional metadata about the data collection. For example, a monitoring location might have two sensors that collect the exact same data but in different locations (e.g, on both sides of a sluice gate). These would have different timeseries IDs.

Note that the `geometry` column contains spatial point data for each row.
We'll be pulling site metadata like this later on, so we can set `skip_geometry` to cut down on the response size (saves us memory, saves the API folks money).
We could set the `properties` argument to further whittle the response down, if we want.

```{r}
#| label: gw-timeseries-ids
# 1. Pull timeseries IDs with POR >= 7 years and obs within last 406 days
gw_all_ts_ids <- read_waterdata_ts_meta(
  begin = paste0("1700-01-01/", latest_date_por),
  end = paste0(earliest_date_needs_observation, "/.."),
  parameter_code = gw_pcodes,
  skipGeometry = TRUE
) |>
  select(
    monitoring_location_id,
    parameter_code,
    statistic_id,
    begin,
    end,
    time_series_id,
    everything()
  ) |>
  arrange(monitoring_location_id, parameter_code, statistic_id)

# 2. daily data are going to be the easiest to work with. We should first figure
# out which sites have enough daily data before looking at alternatives
ml_ids_daily_means <-
  gw_all_ts_ids |>
  filter(statistic_id == "00003")

# 3. some sites might not have daily data but *do* have continuous.
ml_ids_continuous <-
  gw_all_ts_ids |>
  filter(statistic_id == "00011")
```

-------------------------------

The code below starts to get into the weeds of assessing data coverage for Groundwater sites.
This is just based on my own experimentation, but you're welcome to ignore this code and write your own.

```{r}
#| label: gw-field-measurements
#| eval: false

# this cell isn't evaluated by-default as it takes a long time to run. It could be useful for filling data gaps, though.
# 4. Also pull from field-measurements, which includes different sites.
# First fetch sub-list of sites with at least one obs in last 406 days
gw_recent_field_measurements <-
  read_waterdata_field_measurements(
    parameter_code = gw_pcodes,
    time = paste0(earliest_date_por, "/..")
  )

gw_ml_ids_field_measurement <- unique(
  gw_recent_field_measurements$monitoring_location_id
)

# Among sites with at least one obs in last 406 days, determine which have a
# long enough POR
gw_old_field_measurements <-
  # NOTE: parallelized here
  furrr::future_map_dfr(
    # purrr::map_dfr(
    split(
      unique(gw_recent_field_measurements$monitoring_location_id),
      ceiling(
        seq_along(unique(gw_recent_field_measurements$monitoring_location_id)) /
          100
      )
    ),
    ~ read_waterdata_field_measurements(
      monitoring_location_id = .x,
      parameter_code = gw_pcodes,
      skipGeometry = TRUE
    )
  )

gw_field_measurements_sufficient_por <-
  gw_old_field_measurements |>
  sf::st_drop_geometry() |>
  filter(time <= latest_date_por) |>
  distinct(monitoring_location_id) |>
  pull(monitoring_location_id)

gw_old_field_measurements |>
  sf::st_drop_geometry() |>
  filter(monitoring_location_id %in% gw_field_measurements_sufficient_por) |>
  mutate(month = lubridate::floor_date(time, unit = "months")) |>
  select(monitoring_location_id, parameter_code, month) |>
  tidyr::nest(month)
```

Here's where things get more complex.
For each site, we want to determine whether there is at least one observation per month across its entire POR.
Ideally, these observations would be from /daily, as these are going to be the easiest to fetch in the future (easier than /continuous).
So first we fetch the entire daily POR for sites (below is an example of only pulling 50 sites) and determine whether there is at least one observation per month.

Some sites might have data missing for specific months.
Instead of exlcuding these sites, we might consider if we can fill in the data gaps with data from /field-measurements or (as a last resort) /continuous.

```{r}
#| label: gw-compute-frequency
# 5. Assess frequency of observations -- is there at least one obs per timestep
# over POR?

# Fetch daily data since earliest_date_por -- requires splitting TS IDs into
# chunks of 50
all_daily_obs <-
  purrr::map_dfr(
    split(
      ml_ids_daily_means$time_series_id,
      ceiling(seq_along(ml_ids_daily_means$time_series_id) / 50)
    )[1], # NOTE: testing with one chunk of 50 IDs
    ~ read_waterdata_daily(
      time_series_id = .x,
      time = paste0(earliest_date_por, "/.."),
      skipGeometry = TRUE,
      limit = 50000
    )
  )

# Determine if there are any months of missing daily data for each monitoring
# location
daily_missing_timesteps <-
  all_daily_obs |>
  mutate(month = lubridate::floor_date(time, unit = update_timestep)) |>
  arrange(time_series_id, month) |>
  distinct(
    monitoring_location_id,
    parameter_code,
    statistic_id,
    time_series_id,
    month
  ) |>
  group_by(monitoring_location_id, parameter_code) |>
  summarize(
    missing_timesteps = list(required_dates[!required_dates %in% month])
  ) |>
  left_join(
    distinct(ml_ids_daily_means, monitoring_location_id, parameter_code, begin),
    by = c("monitoring_location_id", "parameter_code")
  ) |>
  # only want to track missing months since start of POR ("begin") for each site:
  mutate(
    missing_timesteps = purrr::map2(missing_timesteps, begin, ~ .x[.x >= .y])
  ) |>
  select(-begin) |>
  # only keep ML IDs with missing daily data
  filter(purrr::map_lgl(missing_timesteps, ~ length(.x) > 0))


# These are sites with no gaps in daily data for their entire POR -- these will
# be easiest to pull data for
gw_ml_ids_complete_daily <-
  all_daily_obs |>
  distinct(monitoring_location_id, parameter_code, time_series_id) |>
  anti_join(
    daily_missing_timesteps,
    by = c("monitoring_location_id", "parameter_code")
  )

# These are sites that have gaps in daily data that *may* be filled with
# continuous data, we'll check below
gw_ml_ids_partial_daily <-
  ml_ids_continuous |>
  select(monitoring_location_id, parameter_code, time_series_id, begin) |>
  right_join(
    daily_missing_timesteps,
    by = c("monitoring_location_id", "parameter_code")
  ) |>
  # check if missing daily data dates are before a site started collection
  # continuous data. If yes, then there's really no hope to fill the daily data
  # gaps.
  filter(purrr::map2_lgl(missing_timesteps, begin, ~ all(.x > .y)))


# These are sites with daily data gaps that can't be filled with continuous data
# (daily data gaps occur before continuous measurement began). We can't do
# anything with these. NOTE: maybe we should check /field-measurements data too
gw_ml_ids_lost_causes <- all_daily_obs |>
  distinct(monitoring_location_id, parameter_code, time_series_id) |>
  anti_join(
    bind_rows(gw_ml_ids_complete_daily, gw_ml_ids_partial_daily),
    by = c("monitoring_location_id", "parameter_code")
  )

gw_ml_ids_lost_causes
```

```{r}
#| label: gw-metadata
# Fetch metadata for each monitoring location -- requires splitting list of ML
# IDs into chunks of 100
gw_ml_ids <- unique(c(ml_ids_daily_means$monitoring_location_id, ml_ids_continuous$monitoring_location_id))

gw_ml_metadata <-
  purrr::map_dfr(
    split(gw_ml_ids, ceiling(seq_along(gw_ml_ids) / 100)),
    read_waterdata_monitoring_location
  ) |>
  left_join(
    bind_rows(ml_ids_daily_means, ml_ids_continuous) |>
      select(monitoring_location_id, parameter_code, statistic_id),
    by = "monitoring_location_id"
  )

# Plot ML IDs by parameter code
{
  gw_ml_metadata |>
    ggplot() +
    geom_sf(aes(colour = parameter_code))
} |>
  plotly::ggplotly()
```

:::

